Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-25 paper/paper.tex
mention smallness in the changes
fill in smallness figures
remove todo

We present a selection of the non-linear type constructions in
\Cref{fig:non-linear-types} and provide the rest in
\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of the appendix.  First,
we include universe types $U$ of small non-linear types and $L$ of linear
types. These are defined as
universes ``ala Coquand'' in that we define judgments saying when
non-linear and linear types are \emph{small} and define the universes to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. The definition of smallness is simply that all types are small as long as their sub-formulae are, with the exception of the two universe types themselves. These universe
types are needed so that we can define non-linear and linear types by recursion on natural
numbers.
Next, we include standard $\Sigma,\Pi$, empty, unit, Boolean, and natural number types. More complex inductive types --- such as sum types, list types, and $Fin~n$ --- can be defined in terms of these primitives, and we give their encoding in \cref{TODO} of the appendix.

@@

%% \pagebreak

@@

\ifappendix
\appendix
\section{Syntax}
\begin{figure}
\begin{minipage}[t]{.35\textwidth}
  \begin{align*}
    \el\left(\Var\,M\right) B &= B\,M\\
    \el\left(\mathsf{K}\,A\right) B &= A\\
    \el\left(\bigoplus A\right) B &= \bigoplus_{y:Y}\el(A y)B\\
    \el\left(\bigamp A\right) B &= \bigamp_{y:Y}\el(A y)B\\
    \el\left(A \otimes A'\right) B &= \el(A)B \otimes \el(A')B
  \end{align*}
\end{minipage}%
\begin{minipage}[t]{.65\textwidth}
\begin{align*}
    \map\left(\Var\,M\right)\,f &= f\,M\\
    \map\left(\K\,A\right)\,f &= \lambda a. a\\
    \map\left(\bigoplus A\right)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
    \map\left(\bigwith\,A\right)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
    \map\left(A \otimes A'\right)\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
  \end{align*}
\end{minipage}%
  \caption{Strictly positive functors functorial actions}
  \label{fig:spf-act}
\end{figure}

@@

\begin{align*}
  Fin~0 &= \bot \\
  Fin~(suc n) &= 1 + (Fin~n)
\end{align*}

@@

\begin{figure}
  \begin{mathpar}
    \inferrule{\Gamma \vdash M : U}{\Gamma \vdash \unquoteTy M \isSmall}

    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash Y \isSmall}{\Gamma \vdash \PiTyLimit {x}{X}{Y} \isSmall}
    \cdots
  \end{mathpar}
  \caption{Small Non-linear Types}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule{\Gamma \vdash M : L}{\Gamma \vdash \unquoteTy M \isSmallLin}

    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash A \isSmallLin}{\Gamma \vdash \LinPiTyLimit {x}{X}{A} \isSmallLin}
    \cdots
  \end{mathpar}
  \caption{Small Linear Types}
\end{figure}

>>>>

We present a selection of the non-linear type constructions in
\Cref{fig:non-linear-types} and provide the rest in
\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of the appendix.  First,
we include universe types $U$ of small non-linear types and $L$ of linear
types. These are defined as
universes ``ala Coquand'' in that we define judgments saying when
non-linear and linear types are \emph{small} and define the universes to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. The definition of smallness is simply that all types are small as long as their sub-formulae are, with the exception of the two universe types themselves. These universe
types are needed so that we can define non-linear and linear types by recursion on natural
numbers.
Next, we include standard $\Sigma,\Pi$, empty, unit, Boolean, and natural number
types. More complex inductive types --- such as sum types, list types, and
$Fin~n$ --- can be defined in terms of these primitives, and we give their
encoding in the appendix.

@@

\newpage

@@

\newpage
\ifappendix
\appendix
\section{Syntax}
\begin{figure}
\begin{minipage}[t]{.35\textwidth}
  \begin{align*}
    \el\left(\Var\,M\right) B &= B\,M\\
    \el\left(\mathsf{K}\,A\right) B &= A\\
    \el\left(\bigoplus A\right) B &= \bigoplus_{y:Y}\el(A y)B\\
    \el\left(\bigamp A\right) B &= \bigamp_{y:Y}\el(A y)B\\
    \el\left(A \otimes A'\right) B &= \el(A)B \otimes \el(A')B
  \end{align*}
\end{minipage}%
\begin{minipage}[t]{.65\textwidth}
\begin{align*}
    \map\left(\Var\,M\right)\,f &= f\,M\\
    \map\left(\K\,A\right)\,f &= \lambda a. a\\
    \map\left(\bigoplus A\right)\,f &= \lambda a. \letin{\oplusinj y {a_y}}{a}{\oplusinj y {\map(A\,y)\,f\,a_y}}\\
    \map\left(\bigwith\,A\right)\,f &= \lambda a. \withlamb{y}{\map(A\,y)\,f\,(\withprj y a)}\\
    \map\left(A \otimes A'\right)\,f &= \lambda b. \letin{(a,a')}{b}{(\map(A)\,f\,a,\map(A')\,f\,a')}
  \end{align*}
\end{minipage}%
  \caption{Strictly positive functors functorial actions}
  \label{fig:spf-act}
\end{figure}

@@

\begin{align*}
  Fin~0 &= \bot \\
  Fin~(suc~n) &= 1 + (Fin~n)
\end{align*}

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \inferrule{~}{\Gamma \vdash 1 \isSmall}

    \inferrule{~}{\Gamma \vdash Bool \isSmall}

    \inferrule{~}{\Gamma \vdash \bot \isSmall}

    \inferrule{~}{\Gamma \vdash Nat \isSmall}

    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash Y \isSmall}{\Gamma \vdash \PiTyLimit {x}{X}{Y} \isSmall}

    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash Y \isSmall}{\Gamma \vdash \SigTyLimit {x}{X}{Y} \isSmall}

    \inferrule{\Gamma \vdash M : U}{\Gamma \vdash \unquoteTy M \isSmall}

    \inferrule{\Gamma \vdash A \isSmallLin}{\Gamma \vdash {\ltonl A} \isSmall}

    \inferrule{\Gamma \vdash X \isSmall \and \nonlinterm \Gamma M X \and
      \nonlinterm \Gamma N X}{\Gamma \vdash {M =_{X} N} \isSmall}
  \end{mathpar}
  \caption{Small Non-linear Types}
\end{figure}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \inferrule{~}{\Gamma \vdash I \isSmallLin}

    \inferrule{c \in \Sigma}{\Gamma \vdash {\literal{c}} \isSmallLin}

    \inferrule{\Gamma \vdash A \isSmallLin \and \Gamma \vdash B \isSmallLin}{
      \Gamma \vdash {A \otimes B} \isSmallLin}

    \inferrule{\Gamma \vdash A \isSmallLin \and \Gamma \vdash B \isSmallLin}{
      \Gamma \vdash {A \lto B} \isSmallLin}

    \inferrule{\Gamma \vdash A \isSmallLin \and \Gamma \vdash B \isSmallLin}{
      \Gamma \vdash {A \tol B} \isSmallLin}

    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash A \isSmallLin}{\Gamma \vdash \LinSigTyLimit {x}{X}{A} \isSmallLin}

    \inferrule{\Gamma \vdash X \isSmall \and \Gamma,x:X \vdash A \isSmallLin}{\Gamma \vdash \LinPiTyLimit {x}{X}{A} \isSmallLin}

    \inferrule{\Gamma \vdash A \isSmallLin \and
      \Gamma \vdash B \isSmallLin \and
      \nonlinterm \Gamma f {\ltonl{(A \lto B)}} \and
      \nonlinterm \Gamma g {\ltonl{(A \lto B)}}}{\Gamma \vdash {\equalizer
        {a}{f}{g}} \isSmallLin}

    \inferrule{\Gamma \vdash M : L}{\Gamma \vdash \unquoteTy M \isSmallLin}
  \end{mathpar}
  \caption{Small Linear Types}
\end{figure}
